
\thispagestyle{empty}

\vspace*{0.1in}

\begin{centering}
{\LARGE \NAMD\ Version \NAMDVER}\\
{\large
\bigskip
Authors: M. Bhandarkar, R. Brunner, A. Dalke,\\
A. Gursoy, W. Humphrey, N. Krawetz,\\
M. Nelson, J. Phillips, A. Shinozaki\\
\medskip
Theoretical Biophysics Group, Beckman Institute, University of Illinois.\\
\bigskip
\copyright 1995-99 The Board of Trustees of the University of Illinois.
All Rights Reserved\\
\bigskip
}
\end{centering}

  \rule{6in}{0.04in}				\\	\vspace{0.25in}

{\bf \LARGE \noindent NOTICE}
\vspace{0.25 in}

\noindent The program \NAMD\ is {\it not\/} in the public domain.
However, it is freely available without fee for
education, research, and commercial purposes.  By obtaining copies
of this and other files that comprise the \NAMD\ program, you, 
the Licensee, agree to abide by the following
conditions and understandings with respect to the copyrighted software:

\begin{enumerate}
\item The software is copyrighted in the name of the Board of Trustees
of the University of Illinois (UI), and ownership of the software
remains with the UI.

\item Permission to use and modify this software and its documentation
is hereby granted  to Licensee.  In addition,
permission to copy this work and any derived works is
granted to Licensee, provided that

\begin{enumerate}
\item the copyright notice and this permission notice appear on
all such copies, 
\item that proper credit be given by citing

\begin{quotation}
         \noindent M. Nelson, W. Humphrey, A. Gursoy, A. Dalke,
	 L. Kale, R. Skeel and K. Schulten.
	 NAMD - A parallel, object-oriented molecular dynamics program.
	 {\it J. Supercomputing App.}, 10:251-268, 1996.
\end{quotation}

\noindent for \NAMD\ and

\begin{quotation}
	 \noindent W. T. Rankin and J. A. Board, Jr., Tech Rept. 95-002,
	 EE Dept., Duke Univ.
\end{quotation}

\noindent for DPMTA
\item that \NAMD\ and any derived works are not redistributed for a fee,
\item that programs derived from \NAMD\ be given a different name.
\end{enumerate}

\item Licensee may not use the name, logo, or any other symbol of the UI
    nor the names of any of its employees nor any adaptation thereof in
    advertizing or publicity pertaining to the software without specific
    prior written approval of the UI.

\item THE UI MAKES NO REPRESENTATIONS ABOUT THE SUITABILITY OF THE
    SOFTWARE FOR ANY PURPOSE.  IT IS PROVIDED "AS IS" WITHOUT EXPRESS
    OR IMPLIED WARRANTY.

\item The UI shall not be liable for any damages suffered by Licensee from
    the use of this software.
\end{enumerate}

